### 2017-06-25 paper.tex
methods

%units
\usepackage{siunitx}
\DeclareSIUnit\Molar{\textsc{m}}
\sisetup{scientific-notation = true}
%% Include all macros below

@@

% % Use the asterisk to denote corresponding authorship and provide email address in note below.
% * correspondingauthor@institute.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons and understanding how this occurs has been the focus of many modeling studies. However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature and complex activation kinetics. To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion of the number of species. With this new method, spatial and temporal aspects of the signaling network can be studied without compromising biochemical details. We used this new method to examine how calmodulin molecules, both partially loaded and fully loaded with Ca\textsuperscript{2+}, choose pathways to interact with and activate CaMKII under various Ca\textsuperscript{2+} input conditions. We found that the frequency dependence of the CaMKII phosphorylation with Ca\textsuperscript{2+} input is not primarily determined by subunit number, but is intrinsic to the network and can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion. Depending on whether Ca\textsuperscript{2+} input is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns. The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca2+ channel nanodomains and cytoplasm.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}

Ca\textsuperscript{2+} signals are commonly used by cells for various types of activities. In neurons, Ca\textsuperscript{2+} can regulate gene expression and dendritic spine enlargement, strengthen synaptic connectivity, and promote neural growth or even death. These strikingly contrasting cell processes are possible because Ca\textsuperscript{2+} signals of varying strengths and temporal patterns are interpreted differently by intracellular signaling pathways. In our study, we focus on one particular pathway that involves calmodulin and Calcium/calmodulin-dependent protein kinase II (CaMKII). We have developed a new computational method that gets around the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods. Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca2+ input conditions, calmodulin availability and the Ca2+ diffusion rate.

\linenumbers
% Use "Eq" instead of "Equation" for equation citations.
% Camkii binding with F-actin is mostly reported in spines studies, but also seen in cytoplasm as Li Boxing 2016
%Bhalla's paper CaMKII model with diffusion, but simplified CaM states;
% setup with spine geometry; camca4 only; no diffusion effect; stochastic? hybrid? deterministic
\section*{Introduction}
Calcium/calmodulin-dependent protein kinase II (CaMKII) is an important enzyme widely distributed in the central nervous system and other tissues including cardiac muscles~\cite{Wayman:2008gla,Lisman:2002ki,Erickson:2014fs}. CaMKII is involved in intracellular Ca\textsuperscript{2+} signaling as an effector of the Ca\textsuperscript{2+} sensor protein calmodulin (CaM)~\cite{Herring:2016bh,Hell:2014bd,Coultrap:2012ip}. In neurons, CaMKII molecules are especially involved in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}. In spines, activated CaMKII molecules interact with postsynaptic density proteins, facilitating actins to reorganize, leading to spine enlargement and upregulation of $alpha$-amino-3-hydroxy-5-methyl-4-isoxazolepropionic acid (AMPA) receptor numbers~\cite{Herring:2016bh}. CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}. In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus. Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors. Therefore CaMKII molecules play a key role in excitation-transcription coupling.

The CaMKII molecule is a holoenzyme that consists of 12 subunits. Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase. Activities such as CaM binding and auto-phosphorylation result in various subunits states. For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site. Once phosphorylated, the CaM unbinding rate decreases dramatcially, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}. Even when the CaM molecule unbinds, the CaMKII subunit stays activated, entering an autonomous state. Finally, Thr305/Thr306 can also be phosphorylated. This theronine site overlaps with the CaM binding site and its phosphorylation blocks the binding of CaM molecules. In this case, the subunit becomes "capped".

@@

Another complication is related to the intracellular environment. Conventionally, biochemical reactions are modeled in a deterministic manner using Ordinary Differential Equations (ODE). The underlying assumptions are that reactions occur in a spatially homogenous environment, reactants are abundant and not subject to stochasticity, and molecules diffuse sufficiently fast. The majority of previous modeling studies belong to this category except for a few that are stochastic or hybrid models ~\cite{Bhalla:2004cu, Zeng:2010bq, Holmes:2000uk}. However, intracellular space is highly heterogeneous and compartmentalized~\cite{LubyPhelps:2000uj,Dix:2008gy}. Reactions are often restricted to a small space. Structures such as the cytoskeleton, scaffold proteins or endoplasmic reticulum often act as diffusion barriers to slow down molecules. In addition, most previous modeling studies focus on the dendritic spine, where NMDA receptors-channels are the main Ca\textsuperscript{2+} providers.

In the present work, we focuse on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which servers for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus. Specifically, we examined the Ca\textsuperscript{2+}-CaM-CaMKII network activities near L-type Ca\textsuperscript{2+} channels. This is an area less studied previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}. To study this Ca\textsuperscript{2+}-CaM-CaMKII signaling network and deal with the problem of combinatorial explosion, we modified the published and freely available simulator Smoldyn~\cite{Andrews:2004fs}. Smoldyn is a particle-based stochastic simulator and has been used for simulating reaction and diffusion processes in cells. It works well for relatively simple reaction networks but not for holoenzymes such as CaMKII. We added new data structures to describe the reaction network in a compact way. CaMKII holoenzymes are modeled as a collection of subunits. Each subunit has a set of binding sites. Subunits react independently and diffuse collectively. Reactions are defined between binding sites. The reaction network is stored in a hashtable to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network. We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}. Moreover, we added detailed components to the network and examined the frequency dependence of CaMKII activation. We found that the number of subunits does not play a significant role in CaMKII activation. Instead, slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation. In the presence of a limited number of CaM molecules, CaM must make an intricate choice between binding Ca\textsuperscript{2+} first or a CaMKII subunit first. Interestingly, a change in the decision pattern leads to an altered frequency dependence of the network.

% % Results and Discussion can be combined.
\section*{Results}
\subsection*{Testing of the simulator modifications}
We first tested the modifications to Smoldyn used in this paper (see Methods) using a Ca\textsuperscript{2+}-CaM network (Fig~\ref{fig1}A). The 4 Ca\textsuperscript{2+} binding sites on CaM gives rise to a total of 9 different binding states of a CaM molecule. For the reaction volume we use a $500\times500\times500$ \SI{}{\cubic\nm} cube with all sides being reflective. The size of this volume mimics that of a Ca\textsuperscript{2+} channel nanodomain or a small dendritic spine head. Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.8671}{\micro\Molar} and \SI{9.3023}{\micro\Molar} respectively. The concentrations are chosen to produce an observable amount of 4-Ca\textsuperscript{2+} bound CaM at the steady state. All molecules are initially uniformly distributed. We tested reactions using two different Ca\textsuperscript{2+} diffusion constants \SI{0.0000022}{\square\cm\per\s}~\cite{Keller:2008ez} and half of this value \SI{0.0000011}{\square\cm\per\s}.

We first characterized the reactions in the network to see if results of our stochastic model could reasonably be compared to results with standard ODE methods. A second order chemical reaction in the solvent phase consists of two steps: molecules encountering each other by diffusion followed by the molecules reacting with each other. If the encountering step takes a much longer time to occur than the reacting step, the reaction is diffusion-limited; otherwise, the reaction is activation-limited. Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{},
\begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx \k_{on}$ as diffusion is sufficiently fast. Given the diffusion coefficients for Ca\textsuperscript{2+} and calmodulin and the parameter values for Ca\textsuperscript{2+} and calmodutlin reaction kinetics in Table 1., we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime. This was also true if diffusion constants were reduced 10 fold. ? Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.

We coded a deterministic ODE system in COPASI~\cite{} and compared results with our stochastic model using the same kinetics. We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{2}. Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used. These results suggest that our modifications of Smoldyn are working properly.

\begin{figure}[!h]
	\caption{{\bf}
	A: In this example, N1 and N2 are the two sites of N-lobe; C1 and C2 are the two sites of C-lobe. The binding and unbinding rate constants are labeled at the end of reactions. Between lobes reactions are independent. Within each lobe bindings are cooperative. Thus N2 binding cannot happen without N1 being bound; a similar rule applies to C sites. ‘+’ represents a binding reaction; ‘~’ represents molecules that are bound. ‘{}’ on the left-hand side specifies the binding sites states of reactants or the conditions for a reaction to occur. On the right-hand side, binding sites involved in the reaction are assigned to new values to change state. Notice the ‘==‘ sign on the left-hand side, and the ‘=‘ sign on the right-hand side are different. The ‘==‘ represents True or False of equality, whereas the ‘=‘ denotes an assigning operation. B: Ca\textsuperscript{2+} and CaM interactions, resulting in 9 different states of CaM. Arrows represent the direction of binding reactions. C: The complete Ca\textsuperscript{2+}-CaM-CaMKII network. The layer in blue represents Ca\textsuperscript{2+}-CaM interactions the same as in figure B. Edges are labeled to indicate binding reactions. For example, N1 means binding of a Ca\textsuperscript{2+} ion on CaM N1 site. KN1 means CaM attached to a CaMKII subunit binds a Ca\textsuperscript{2+} ion on N1 site. KNxCy means CaMKII subunit binds a CaM has $x$ Ca\textsuperscript{2+} ions bound on N-lobe and $y$ Ca2+ on C-lobe, $x$ and $y$ are from 0 to 2. KpNxCy means phosphorylated CaMKII subunit binds a CaM. Red edges mark the preferred pathway. D: Reaction scheme1. From top to bottom, the blue edges represent autophosphorylations of CaM N0C2, N1C2 and N2C2 respectively. The layer colored yellow represents Ca\textsuperscript{2+}-CaM interactions, the same as in panel B. E. Reaction scheme22. From left to right, the blue edges represent autophosphorylations of CaM N2C0, N2C1 and N2C2 respectively. The front layer(Layer2) colored yellow represents interactions between Ca\textsuperscript{2+} and CaMKII attached CaM molecules. Layer3 represents interactions between Ca\textsuperscript{2+} and phosphorylated CaMKII attached CaM molecules.}
\label{fig1}
\end{figure}

\begin{figure}[!h]
	\caption{{\bf}
	Time courses of 9 states of CaM molecules. The smooth lines are simulated using COPASI, and the discontinous lines are from the simple stochastic models with two different diffusion coefficients.}
\label{fig2}
\end{figure}

@@

We set up a prototype model to identify the major network branches that contribute to the phosphorylation of CaMKII subunits. The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig1}. In this network diagram molecule species are represented as vertices and reactions are represented as the network edges. CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in the $1 \times 1 \times 2$\SI{}{\cubic\um} box. As an initial contion, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII. The initial state is at an equilibrium, which we intentionally computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules (Fig~\ref{fig}). Ca\textsuperscript{2+} ions were released as a single source from the top $1 \times 1$ \SI{}{\square\um}. A previously generated input file having a total of 6 Ca2+ bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods). We ran the simulation up to \SI{2.5}{\s}, recorded the molecule numbers at every \SI{1}{\ms} and logged all reaction events. Using the events log, we counted the accumulated occurrences of each reaction type at every \SI{10}{\ms}, starting from the initial steady-state until the Ca\textsuperscript{2+} bursts were over.

For a particular reaction, the number of occurrences during a time span is considered to be the net number of molecule state changes along the corresponding edge, i.e., the number of unbinding events is subtracted from the number of binding events. A negative number means that unbinding occurs more often than binding within the given time span.

We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{}). The network consists of three layers (Fig~\ref{}). The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2) as shown separately in Fig~\ref{}. The front layer (Layer2) is for reactions of Ca\textsuperscript{2+} with CaM attached to unphosphorylated CaMKII (KNxCy). The middle layer (Layer3) is for reactions of Ca\textsuperscript{2+} with CaM attached to phosphorylated CaMKII (KpNxCy). Starting in Layer1, there are at most 4 possible ways for each CaM species to change state: binding a Ca\textsuperscript{2+} at a C site, binding Ca\textsuperscript{2+} at an N site, binding to an unphosphorylated CaMKII subunit or binding to a phosphorylated CaMKII subunit. However, binding to a phosphorylated CaMKII subunit requires a bound subunit to be phosphorylated first and then lose the bound CaM. This type of reaction rarely occurs during the early stage of the simulation since phosphorylation is slow and CaM unbinding from a phosphorylated subunit is uncommon. Thus we focus on the first 3 types of reactions.

>>>>

%units
\usepackage{siunitx}
\DeclareSIUnit\Molar{\textsc{m}}
%\sisetup{scientific-notation = true}
%% Include all macros below

@@

% % Use the asterisk to denote corresponding authorship and provide email address in note below.
% * correspondingauthor@institute.edu
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Calcium/calmodulin-dependent protein kinase II (CaMKII) holoenzymes play a critical role in decoding Ca\textsuperscript{2+} signals in neurons and understanding how this occurs has been the focus of many modeling studies. However, CaMKII is notoriously difficult to model and simulate in detail because of its multi-subunit nature, which causes a combinatorial explosion of the number of species that must be modeled. To study the Ca\textsuperscript{2+}-calmodulin-CaMKII reaction network with detailed kinetics while including the effect of diffusion, we have customized an existing stochastic particle-based simulator Smoldyn to manage the problem of combinatorial explosion. With this new method, spatial and temporal aspects of the signaling network can be studied without compromising biochemical details. We used this new method to examine how calmodulin molecules, both partially loaded and fully loaded with Ca\textsuperscript{2+}, choose pathways to interact with and activate CaMKII under various Ca\textsuperscript{2+} input conditions. We found that the frequency dependence of the CaMKII phosphorylation is intrinsic to the network kinetics and the activation pattern can be modulated by the relative amount of Ca\textsuperscript{2+} to calmodulin and by the rate of Ca\textsuperscript{2+} diffusion. Depending on whether Ca\textsuperscript{2+} influx is saturating or not, calmodulin molecules could choose different routes within the network to activate CaMKII subunits, resulting in different frequency dependence patterns. In addition, the size of the holoenzyme shows a subtle effect on CaMKII activation. The more extended the subunits are organized, the easier for CaM molecules to access and activate the subunits. The findings suggest that particular intracellular environmental factors such as crowding and calmodulin availability can play an important role in decoding Ca\textsuperscript{2+} signals and can give rise to distinct CaMKII activation patterns in dendritic spines, Ca\textsuperscript{2+} channel nanodomains and cytoplasm.
% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}

Ca\textsuperscript{2+} signals are commonly used by cells for various types of activities. In neurons, Ca\textsuperscript{2+} can regulate gene expression and dendritic spine enlargement, strengthen synaptic connectivity, and promote neural growth or even death. These strikingly contrasting cell processes are possible because Ca\textsuperscript{2+} signals of varying strengths and temporal patterns are interpreted differently by intracellular signaling pathways. In our study, we focus on one particular pathway that involves calmodulin and Ca\textsuperscript{2+}/calmodulin-dependent protein kinase II (CaMKII). We have developed a new computational method that avoids the problem of combinatorial explosion inherent with modeling this pathway in biophysical detail, and this allows us to model this pathway more realistically than conventional methods. Experiments have shown that CaMKII activity is sensitive to Ca\textsuperscript{2+} signal frequency and our models demonstrate how this frequency dependence depends on Ca\textsuperscript{2+} input conditions, calmodulin availability and the Ca\textsuperscript{2+} diffusion rate.

\linenumbers
% Use "Eq" instead of "Equation" for equation citations.
% Camkii binding with F-actin is mostly reported in spines studies, but also seen in cytoplasm as Li Boxing 2016
%Bhalla's paper CaMKII model with diffusion, but simplified CaM states;
% setup with spine geometry; camca4 only; no diffusion effect; stochastic? hybrid? deterministic
\section*{Introduction}
Calcium/calmodulin-dependent protein kinase II (CaMKII) is an important enzyme widely distributed in the central nervous system and other tissues including cardiac muscles~\cite{Wayman:2008gla,Lisman:2002ki,Erickson:2014fs}. CaMKII is involved in intracellular Ca\textsuperscript{2+} signaling as an effector of the Ca\textsuperscript{2+} sensor protein calmodulin (CaM)~\cite{Herring:2016bh,Hell:2014bd,Coultrap:2012ip}. In neurons, CaMKII molecules are essential in long-term potentiation (LTP), which is our best model for the molecular basis of learning and memory~\cite{Bliss:1973jg}. In spines, activated CaMKII molecules interact with postsynaptic density proteins, facilitating actins to reorganize, leading to spine enlargement and upregulation of $alpha$-amino-3-hydroxy-5-methyl-4-isoxazolepropionic acid (AMPA) receptor numbers~\cite{Herring:2016bh}. CaMKII molecules can also phosphorylate AMPA receptors to regulate channel conductance~\cite{Herring:2016bh,Coultrap:2012ip} or form complexes with NMDA receptors~\cite{Hell:2014bd}. In addition to synaptic functions, recent studies~\cite{Ma:2014dr,Li:2016cq} suggest that CaMKII in the soma can act as carriers to shuttle Ca\textsuperscript{2+}-CaM into the nucleus. Ca\textsuperscript{2+}-CaM molecules are then unloaded in the nucleus, participate in the calcium/calmodulin-dependent protein kinase IV (CaMKIV) cascade to activate nuclear transcription factors. Therefore CaMKII molecules play a key role in excitation-transcription coupling.

The CaMKII molecule is a holoenzyme that consists of 12 subunits grouped in 2 rings. Each subunit contains an association domain allowing the formation of multimers, a regulatory domain with CaM binding sites and phosphorylation sites, and a catalytic domain to act as a kinase. Activities such as CaM binding and auto-phosphorylation result in various subunits states. For example, in the presence of Ca\textsuperscript{2+}-CaM complexes, each subunit can bind a CaM molecule and expose its catalytic domain to phosphorylate the direct neighbor subunit at its Thr286/Thr287 site. Once phosphorylated, the CaM unbinding rate decreases dramatcially, leading to a prolonged activation of the subunit; we say, the CaM molecule is "trapped" by the CaMKII subunit~\cite{Meyer:1992dp}. Even when the CaM molecule unbinds, the CaMKII subunit stays activated, entering an autonomous state. Finally, Thr305/Thr306 can also be phosphorylated. There theronine sites overlap with the CaM binding site and its phosphorylation blocks the binding of CaM molecules. In this case, the subunit becomes "capped".

@@

Another complication is related to the intracellular environment. Conventionally, biochemical reactions are modeled in a deterministic manner using Ordinary Differential Equations (ODEs). The underlying assumptions are that reactions occur in a spatially homogenous environment, reactants are abundant and not subject to stochasticity, and molecules diffuse sufficiently fast. The majority of previous modeling studies belong to this category except for a few that are stochastic or hybrid models ~\cite{Bhalla:2004cu, Zeng:2010bq, Holmes:2000uk}. However, intracellular space is highly heterogeneous and compartmentalized~\cite{LubyPhelps:2000uj,Dix:2008gy}. Reactions are often restricted to a small space. Structures such as the cytoskeleton, scaffold proteins or endoplasmic reticulum often act as diffusion barriers to slow down molecules. In addition, most previous modeling studies focus on the dendritic spine, where NMDA receptor-channels are the main Ca\textsuperscript{2+} providers.

In the present work, we focus on the soma where voltage-dependent Ca\textsuperscript{2+} channels (CaV) provide the Ca\textsuperscript{2+} influx, which serves for our future study on modeling Ca\textsuperscript{2+}-CaM transportation from surface to the nucleus. Specifically, we examined Ca\textsuperscript{2+}-CaM-CaMKII network activity near L-type Ca\textsuperscript{2+} channels. This is an area less studied previous models, but nonetheless plays a critical role in excitation-transcription coupling~\cite{Ma:2015bg,Li:2016cq}. To study this Ca\textsuperscript{2+}-CaM-CaMKII signaling network and deal with the problem of combinatorial explosion, we modified the published and freely available simulator Smoldyn~\cite{Andrews:2004fs}. Smoldyn is a particle-based stochastic simulator and has been used for simulating reaction and diffusion processes in cells. It works well for relatively simple reaction networks but not for holoenzymes such as CaMKII. We added new data structures to describe the reaction network in a compact way. CaMKII holoenzymes are modeled as a collection of subunits. Each subunit has a set of binding sites. Subunits react independently and diffuse collectively. Reactions are defined between binding sites. The reaction network is stored in a hashtable to allow for lookup during simulations only when reactants collide, without expanding and loading a complete network. We tested and verified these modifications by modeling a Ca\textsuperscript{2+}-CaM interaction network and comparing results with COPASI~\cite{Hoops:2006gy}. Moreover, we added detailed components to the network and examined the frequency dependence of CaMKII activation. We found that slow diffusion of Ca\textsuperscript{2+} can dramatically boost CaMKII activation. Also in the presence of a limited number of CaM molecules, CaM must make an intricate choice between binding Ca\textsuperscript{2+} first or a CaMKII subunit first. Interestingly, a change in the decision pattern leads to an altered frequency dependence of the network.

% % Results and Discussion can be combined.
\section*{Results}
\subsection*{Testing of the simulator modifications}
We first tested the modifications to Smoldyn used in this paper (see Methods) using a Ca\textsuperscript{2+}-CaM network (Fig~\ref{fig1}A). The 4 Ca\textsuperscript{2+} binding sites on CaM gives rise to a total of 9 different binding states of a CaM molecule. For the reaction volume we use a $\SI{500}{\nm}\times\SI{500}{\nm}\times\SI{500}{\nm}$ cube with all sides being reflective. The size of this volume mimics that of a Ca\textsuperscript{2+} channel nanodomain or a small dendritic spine head. Initially, the cube contains 3000 Ca\textsuperscript{2+} ions and 700 apoCaM molecules, equivalent to \SI{39.867}{\micro\Molar} and 9.302\SI{}{\micro\Molar} respectively. The concentrations are chosen to produce an observable amount of 4-Ca\textsuperscript{2+} bound CaM at the steady state. All molecules are initially uniformly distributed. We tested reactions using two different Ca\textsuperscript{2+} diffusion constants \SI{2.2e-6}{\square\cm\per\s}~\cite{Keller:2008ez} and half of this value \SI{1.1e-6}{\square\cm\per\s}.

We first characterized the reactions in the network to see if results of our stochastic model could reasonably be compared to results with standard ODE methods. A second order chemical reaction in the solvent phase consists of two steps: molecules encountering each other by diffusion followed by the molecules reacting with each other. If the encountering step takes a much longer time to occur than the reacting step, the reaction is diffusion-limited; otherwise, the reaction is activation-limited. Conventionally, an experimentally measured binding kinetic rate, $k_{on}$, can be decomposed into an encounter rate, $k_{enc}$, and an intrinsic activation rate, $k_{a}$, using the following equation~\cite{Bamford198505},
\begin{equation}\frac{1}{k_{on}}=\frac{1}{k_{enc}}+\frac{1}{k_a}\end{equation}
For diffusion-limited reactions, $k_a \gg k_{on}$ and $k_{on} \approx k_{enc}$; for activation-limited reactions, $k_a \approx k_{on}$ as diffusion is sufficiently fast. Given the diffusion coefficients for Ca\textsuperscript{2+} and CaM and the parameter values for Ca\textsuperscript{2+} and CaM reaction kinetics in Table S2, we calculated the $\frac{k_a}{k_{on}}$ ratios and concluded that the reactions in the network belong to the activation-limited regime. This was also true if diffusion constants were reduced 10 fold. Thus this simple model resembles a well-mixed system and a standard ODE model will provide a good test standard.

We coded a deterministic ODE system in COPASI~\cite{Hoops:2006gy} and compared results with our stochastic model using the same kinetics. We found that the stochastic model and the ODE COPASI model exhibited similar time courses for all CaM binding state species as show in Fig~\ref{fig2}. Slowing down Ca\textsuperscript{2+} diffusion did not significantly alter the time courses for the parameter values used. These results suggest that our modifications of Smoldyn are working properly.

\begin{figure}[!h]
	\caption{{\bf Ca\textsuperscript{2+}-CaM interaction network.}
	A: Ca\textsuperscript{2+} and CaM interactions can produce 9 different states of CaM. CaM has two lobes, an N lobe and a C lobe, each of which has 2 Ca\textsuperscript{2+} binding sites. Binding sites on the lobes are designated N1, N2, C1 and C2. The N lobe and C lobe bind Ca\textsuperscript{2+} independently of each other, but Ca\textsuperscript{2+} binding within a lobe is cooperative. Thus site N2 cannot bind Ca\textsuperscript{2+} unless N1 has already bound Ca\textsuperscript{2+} and a similar rule applies to sites C2 and C1. In the figure N0C0 represents apoCaM (no Ca\textsuperscript{2+} bound), "N2" implies binding sites N1 and N2 are both bound with Ca\textsuperscript{2+} and "C2" implies similarly that both C1 and C2 have Ca\textsuperscript{2+} bound, "N1" and "C1" imply only the "N1" site and only the "C1" site have Ca\textsuperscript{2+} bound. Arrows represent the direction of binding reactions.
	B: The same reactions between Ca\textsuperscript{2+} and CaM as shown in A are represented using the syntax of the modified simulator. The binding and unbinding rate constants are labeled at the end of reactions. ‘+’ represents a binding reaction; ‘$\sim$’ represents molecules that are bound. ‘\{\}’ on the left-hand side specifies the states of binding sites of reactants or the conditions for a reaction to occur. On the right-hand side, binding sites involved in the reaction are assigned to new values. Notice the ‘==‘ sign is used on the left-hand side, but the ‘=‘ sign is used on the right-hand side. The ‘==‘ represents True or False of equality, whereas the ‘=‘ denotes an assigning operation. 
	}
\label{fig1}
\end{figure}

\begin{figure}[!h]
	\caption{{\bf Test of Smoldyn modifications with a simple Ca\textsuperscript{2+}-CaM reaction network.}
	Plots compare the number of CaM molecules in each of the 9 possible states as computed with the modified stochastic simulator and COPASI. Thick smooth lines are results using COPASI and thin black and red lines are results with the modified simulator for two different values of the calcium diffusion coefficient.
	}
\label{fig2}
\end{figure}

As another verification, we tested the steady state fraction of phosphorylation of CaMKII holoenzymes when they are modeled as multi-subunit complex. As Michalski and Loew~\cite{Michalski:2012ds} pointed out, in a closed system, assuming saturating amount of fully loaded CaM molecules are provided and subunit phosphorylations are allowed only from neighbors that are CaM bound but not those phosphorylated, at the steady state, there will be distinctive phosphorylation levels depending on the number of subunits in the holoenzyme. In particular, for dimers the steady state fraction of phosphorylated subunits is $\frac{1}{2}$; for trimers, $\frac{2}{3}$; for large subunit numebrs, the fraction approaches $1-e^{-1}$. We built a model using a closed system with $\SI{1}{\um}\times\SI{1}{\um}\times\SI{1}{\um}$ geomemtry. 6000 CaMKII subunits and 6020 fully loaded CaM were included in the system. Only the following reactions were allowed: bindings and unbindings between CaM and CaMKII, as well as phosphorylations by neighbor subunits that were bound but not phosphorylated. As expected, we obtained steady state phosphorylation levels that match the Michalski and Loew's formula. Finally, using the same closed system, we tested the complete network and compared with results obtained from a simulation implemented with spatial Gillespie algorithm by Zeng and Holmes~\cite{Zeng:2010bq}. The two simulations showed comparable results.

@@

We set up a prototype model to identify the major network branches that contribute to the phosphorylation of CaMKII subunits. The prototype model comprises Ca\textsuperscript{2+}-CaM-CaMKII interactions as shown in Fig~\ref{fig3}A. In this network diagram molecule species are represented as vertices and reactions are represented as the network edges. CaM molecules and CaMKII holoenzymes with two 6-subunit rings were initially uniformly distributed in a $\SI{1}{\um}\times\SI{1}{\um}\times\SI{2}{\um}$ box. As an initial condition, the box contains 6020 apoCaM molecules (\SI{5}{\micro\Molar}) and 12036 CaMKII subunits (\SI{10}{\micro\Molar}), within which 200 CaM molecules are bound to CaMKII. The initial state is at an equilibrium, which we carefully computed by running 5 simulation trials starting with the same numbers of CaM and CaMKII molecules. For simplicity, we do not consider the resting level intracellular Ca\textsuperscript{2+} contributed significantly to the steady state initial condition. Extracellular Ca\textsuperscript{2+} ions were released as a single source from the top center of box ($\SI{1}{\um}\times$\SI{1}{\square\um}). A previously generated input file having a total of 6 Ca\textsuperscript{2+} bursts delivered at 5 Hz was used to provide Ca\textsuperscript{2+} influx during the simulation (see Methods). We ran the simulation up to \SI{2.5}{\s}, recorded the molecule numbers at every \SI{1}{\ms} and logged all reaction events. Using the events log, we counted the accumulated occurrences of each reaction type at every \SI{10}{\ms}, starting from the initial steady-state until the Ca\textsuperscript{2+} bursts were over. For a particular reaction, the number of occurrences during a time span is considered to be the net number of molecule state changes along the corresponding edge, i.e., the number of unbinding events is subtracted from the number of binding events. A negative number means that unbinding occurs more often than binding within the given time span.

\begin{figure}[!h]
	\caption{{\bf The complete Ca\textsuperscript{2+}-CaM-CaMKII network.}
	A: Species are represented by vertices. Edges are labeled to indicate binding reactions. For example, "N1\_rxn" means binding of a Ca\textsuperscript{2+} ion on CaM at the N1 site. "KN1\_rxn" means CaM attached to a CaMKII subunit binds a Ca\textsuperscript{2+} ion on its N1 site. "K-NxCy\_rxn" means a CaMKII subunit binds a CaM that has x Ca\textsuperscript{2+} ions bound on the N-lobe and y Ca\textsuperscript{2+} ions bound on the C-lobe (x and y range from 0 to 2). "KpNxCy\_rxn" means a phosphorylated CaMKII subunit binds a CaM with x and y Ca\textsuperscript{2+} ions bound on the N and C lobes respectively. Red edges mark the predominant pathway. B: Layers shaded with blue colors. The back layer(Layer1) represents interactions between Ca\textsuperscript{2+} and CaM. The front layer(Layer2) represents interactions between Ca\textsuperscript{2+} and CaM attached to a CaMKII subunit. Layer3 represents interactions between Ca\textsuperscript{2+} and CaM attached to a phosphorylated CaMKII subunit.
	}
\label{fig3}
\end{figure}

We analyzed the edges in the Ca\textsuperscript{2+}-CaM-CaMKII network (Fig~\ref{fig3}A). The network consists of three layers (Fig~\ref{fig3}B). The back layer (Layer1) describes interactions between Ca\textsuperscript{2+} and the 9 states of CaM (NxCy, x,y=0,1,2). The front layer (Layer2) is for reactions of Ca\textsuperscript{2+} with CaM attached to unphosphorylated CaMKII (KNxCy). The middle layer (Layer3) is for reactions of Ca\textsuperscript{2+} with CaM attached to phosphorylated CaMKII (KpNxCy). Starting in Layer1, there are at most 4 possible ways for each CaM species to change state: binding a Ca\textsuperscript{2+} at a C site, binding Ca\textsuperscript{2+} at an N site, binding to an unphosphorylated CaMKII subunit or binding to a phosphorylated CaMKII subunit. However, binding to a phosphorylated CaMKII subunit requires a bound subunit to be phosphorylated first and then lose the bound CaM. This type of reaction rarely occurs during the early stage of the simulation since phosphorylation is slow and CaM unbinding from a phosphorylated subunit is uncommon. Thus we focus on the first 3 types of reactions.
